---
title: "title"
author:
   - name: Hana Dauti
     email: hana.dauti@baruchmail.cuny.edu
     affiliations:
      - department: Department of Mathematics
      - name: Baruch College
      - address: 17 Lexington Avenue
      - city: New York
      - region: New York
      - postal_code: 10010
      - country: United States
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 2
    number-sections: true
    code-folding: hide
    theme: readable
    embed-resources: true
  docx:
    toc: true
    number-sections: true
    highlight-style: github
  pdf:
    number-sections: true
    colorlinks: true
    keeptex: true
    include-in-header: 
      text: |
        \usepackage{booktabs}
        \usepackage{siunitx}
        \newcolumntype{d}{S[
            input-open-uncertainty=,
            input-close-uncertainty=,
            parse-numbers = false,
            table-align-text-pre=false,
            table-align-text-post=false
date: 'last-modified'
date-format: '[This version:] MMMM D, YYYY [<br>(First version: February  16, 2023)]'
abstract: 'text'
bibliography: references.bib
csl: marine-ecology-progress-series.csl
tbl-cap-location: top
number-sections: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
editor:
  mode: source
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

THIS TEMPLATE IS DESIGNED TO HELP GUIDE YOUR PROJECT DEVELOPMENT. YOU
CAN CHANGE OR ADD ANY SECTIONS, BUT I NEED A BRIEF BACKGROUND (WHAT IS
YOUR QUESTION, WHY, WHAT DOES THE DATA LOOK LIKE?). THEN YOU NEED TO
IMPORT THE DATA, CHECK IT, PERFORM EDA, RUN AND INTERPRET TESTS, DEVELOP
APPROPRIATE PLOTS, AND THEN BRIEFLY SUMMARIZE YOUR FINDINGS.

This template also introduces the idea of using quarto to write a
reproducible paper. You may need to play with the execute options in R.
Most of this follows a paper [@bauer2023]. Notice this is a citation
input from my zotero (which then saved it in this projects
references.bib file); more information on citations \@
<https://quarto.org/docs/authoring/citations.html>. The paper is now
shown in my references, with a link to the document.

## Background and metadata

Before diving into the goal of this project, it’s important to provide
some context on the sport and why analyzing it is meaningful. Sports
betting is a global industry worth over \$500 billion, and football
(soccer) is at the heart of it — played by over 250 million people
across more than 200 countries, making it the most popular sport in the
world. Among all leagues, the English Premier League (EPL) stands out as
the most-watched and competitive domestic league, drawing fans,
analysts, and bettors from around the globe. This global appeal and the
high stakes involved make the EPL a valuable subject for data analysis
and performance research.

Understanding how player traits relate to value is important in both
real-life scouting and sports analytics. Teams and analysts often rely
on data like this to make decisions about transfers, contracts, and
player development. Although this model simplifies the real world
**heavily** by excluding factors like age, potential, or fame, it still
reveals useful insights—specifically that pace and passing are strong,
statistically significant predictors of value.

The primary goal of this project is to determine how a soccer player's
on-field attributes—such as pace, shooting, and passing affect their
market value. Using statistical modeling, specifically multiple linear
regression, I will attempt to identify which performance-related traits
are most strongly associated with a player's estimated monetary worth in
the transfer market. The data was obtained from the FIFA 21 Complete
Player Dataset on Kaggle, an open-source data platform. This dataset
compiles detailed information on over 18,000 professional players,
including their skill ratings, demographic information, and financial
estimates like market value and wage.

## Read in (Import) your data

.

```{r}
player_data <- read.csv("fifa21 raw data v2.csv", stringsAsFactors = TRUE)
```

## Check your data

```{r}
head(player_data)
```

```{r}
str(player_data)
```

```{r}
summary(player_data)
```

## Exploratory data analysis

```{r}
library(ggplot2)
```

```{r}
library(tidyverse)


# Clean relevant columns
clean_num <- function(x) as.numeric(sub(" .*", "", x))
player_data$PAC <- clean_num(player_data$PAC)
player_data$SHO <- clean_num(player_data$SHO)
player_data$PAS <- clean_num(player_data$PAS)

# 4. Convert Value to numeric (in euros)
convert_value <- function(val) {
  val <- gsub("€", "", val)
  val <- gsub("K", "e3", val)
  val <- gsub("M", "e6", val)
  val <- gsub("B", "e9", val)
  as.numeric(eval(parse(text = val)))
}

player_data$ValueEUR <- sapply(player_data$Value, convert_value)


playerdata_clean <- player_data %>%
  filter(!is.na(PAC), !is.na(SHO), !is.na(PAS), !is.na(ValueEUR))
```

```{r}
library(tidyr)
playerdata_clean<- separate_wider_delim(player_data, Positions,  ",", names=c("primary_position", NA, NA),too_few="align_start")
ggplot(playerdata_clean, aes(x = primary_position, y = ValueEUR)) +
  geom_boxplot(fill = "lightgreen") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Market Value by Position", x = "Position", y = "Market Value (€)") +
  theme(axis.text.x = element_text(angle = 90, size = 8, hjust = 1))

```

```{r}
ggplot(player_data, aes(x = PAC, y = ValueEUR)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Pace vs. Market Value", x = "Pace", y = "Market Value (€)")

```

## Run statistical tests

Null hypothesis H₀: There is no significant relationship between player
attributes (Pace, Passing, and Shooting) and their market value.

Alternative hypothesis H₁: At least one of the player attributes (Pace,
Passing, or Shooting) has a significant effect on market value.

```{r}
value_model <- lm(ValueEUR ~ PAC + SHO + PAS, data = playerdata_clean)

plot(value_model)
```

Assumptions mostly met, not perfectly, but we have to account for the
fact that the bigger the estimate/ variance = more noise, so for the
sake of this report, it will do.

```{r}
summary(value_model)
```

```{r}
value_model2 <- lm(ValueEUR ~ PAC + PAS, data = playerdata_clean)

plot(value_model2)
```

```{r}
summary(value_model2)
```

## Publication-quality graphs

ggplot2 is your friend here! Remember to add in labels for axes (with
units!), a good title, and pick the right graph type! Examples below for
a few common needs using the iris dataset

```{r}
ggplot(player_data, aes(x = ValueEUR)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  scale_x_log10(labels = scales::dollar_format(prefix = "€")) +
  labs(
    title = "Distribution of Player Market Value",
    x = "Market Value (log scale)",
    y = "Number of Players"
  ) +
  theme_minimal()

```

```{r}
# Add predicted values to the data
player_data$Predicted_Value <- predict(value_model2)

# Plot
library(scales)  # for comma format
ggplot(player_data, aes(x = ValueEUR, y = Predicted_Value)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Predicted vs Actual Player Market Value",
    x = "Actual Market Value (€)",
    y = "Predicted Market Value (€)"
  ) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

```{r}
ggplot(player_data, aes(x = PAS, y = ValueEUR)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Effect of Passing on Market Value",
    x = "Passing Score",
    y = "Market Value (€)"
  ) +
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

```{r}
ggplot(player_data, aes(x = PAC, y = ValueEUR, color = PAS)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Player Market Value by Pace and Passing",
    subtitle = "FIFA 21 data — color indicates passing ability",
    x = "Pace Rating",
    y = "Market Value (€)",
    color = "Passing"
  ) +
  theme_minimal()

```

## Brief discussion

This analysis explored how a soccer player's individual
attributes—specifically pace and passing—influence their market value.
The dataset is quite large, containing over 18,000 players and many
variables. Therefore, for the sake of my project I focused on just a few
key attributes to keep the scope manageable and to demonstrate my
ability to apply regression techniques. Using data from the FIFA 21
player dataset and a multiple linear regression model, I found that both
pace and passing were statistically significant predictors of a player's
value, while shooting was not and was removed from the final model.

The final model showed that higher scores in pace and passing are
strongly associated with higher estimated transfer values. This makes
intuitive sense, as these skills are critical across many field
positions and are likely to influence both real-world performance and
perceived potential in the FIFA dataset. This was also made very clear
visually through the graphs used for exploratory data analysis and my
final graphs.

While the model does a good job of identifying general trends, there are
of course limitations. For example, it doesn’t account for factors like
age, club popularity, or player reputation—non-skill traits that can
greatly influence market value in real-world scenarios. Future analyses
could incorporate these variables for a more robust model.

Overall, this project demonstrates how statistical analysis can reveal
the measurable impact of specific skills on player valuation in
professional soccer, and highlights the usefulness of regression
techniques in sports analytics.

## Notes on rendering

When you use the **render** menu button, Rstudio only produces the first
format (here, the html). To get everything, you can go to the the
terminal tab (next to the console tab in bottom left window) and type

```         
quarto render report_template.qmd
```

that will produce all listed formats (here, html, docx, and pdf).
